---
title: "Week 5 - Functions and `purrr`"
author: "Emorie D Beck"
format: 
  revealjs:
    incremental: true
    code-tools: true
    code-copy: true
    code-line-numbers: true
    code-link: true
    preview-links: true
    slide-number: true
    self-contained: true
    fig-height: 4
    fig-width: 6
    fig-align: center
    css: custom.css
    theme: psc290-23
    # highlight-style: atom-one-dark
    margin-left: "0"
    margin-right: "0"
    width: 1400
    # height: 900
    footer: "PSC 203a - Data Management and Cleaning"
    logo: "https://github.com/emoriebeck/psc290-data-viz-2022/raw/main/01-week1-intro/02-code/02-images/ucdavis_logo_blue.png"
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r, echo = F}
library(knitr)
library(psych)
library(lme4)
library(broom)
library(broom.mixed)
library(kableExtra)
library(plyr)
library(tidyverse)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,	
                      warning = FALSE,
                      results = 'show',
                      fig.width = 4, 
                      fig.height = 4, 
                      fig.retina = 3)
options(htmltools.dir.version = FALSE
        , knitr.kable.NA = "")
```

# Outline

1.  Questions on Homework
2.  Functions
3.  `purrr`
4.  Problem Set and Question Time

# Functions {background-color="blue"}

## [Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

-   A **`function`** or subroutine is a sequence of program instructions that performs a specific task, packaged as a unit.
-   Functions typically take some input and return some specified output

## [Why Write Your Own Functions?]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

-   Automate your workflow
-   Prevent copy-paste errors (only need to update the function, not update the same task in a bunch of areas)
-   Saves you time by providing tools that port to new projects
-   Functions make you *think*, which improves the quality of our work

## [When Should You Write Your Own Functions?]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

-   Hadley Wickham's rule of thumb (which is common in a lot of CS and DS circles) is that if you have to copy-paste something more than twice, you should write a function
-   But really you can write one whenever you'd like
-   I often write functions when a line of code starts getting convoluted because it's usually a good signal that I'm trying to do too much at once.

## [Types of Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

-   Vector functions take one or more vectors as input and return a vector as output.
-   Data frame functions take a data frame as input and return a data frame as output.
-   Plot functions that take a data frame as input and return a plot as output.
-   and so on...
-   We'll talk extensively about this again when we talk about building figures and tables into your workflow in Week 8

## [Examples: Vector Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

-   In my work, I often want to POMP (**P**ercentage **O**f **M**aximum **P**ossible) score or z-score (standardize) variables, which allows me to get data on a standard and interpretable scale (percentage or standard deviations)


::: fragment
::::::: columns
:::: {.column width="75%"}

```{r, eval = F}
bfi %>%
  select(matches("1$")) %>%
  mutate(
    E1 = (E1 - min(E1, na.rm = T))/(max(E1, na.rm = T) - min(E1, na.rm = T))*100
    , A1 = (A1 - min(A1, na.rm = T))/(max(A1, na.rm = T) - min(A1, na.rm = T))*100
    , C1 = (C1 - min(C1, na.rm = T))/(max(C1, na.rm = T) - min(C1, na.rm = T))*100
    , N1 = (N1 - min(N1, na.rm = T))/(max(N1, na.rm = T) - min(N1, na.rm = T))*100
    , O1 = (O1 - min(O1, na.rm = T))/(max(O1, na.rm = T) - min(O1, na.rm = T))*100
  ) %>%
  head(10)
```

::::
:::: {.column width="25%"}

```{r, echo = F}
bfi %>%
  select(matches("1$")) %>%
  mutate(
    E1 = (E1 - min(E1, na.rm = T))/(max(E1, na.rm = T) - min(E1, na.rm = T))*100
    , A1 = (A1 - min(A1, na.rm = T))/(max(A1, na.rm = T) - min(A1, na.rm = T))*100
    , C1 = (C1 - min(C1, na.rm = T))/(max(C1, na.rm = T) - min(C1, na.rm = T))*100
    , N1 = (N1 - min(N1, na.rm = T))/(max(N1, na.rm = T) - min(N1, na.rm = T))*100
    , O1 = (O1 - min(O1, na.rm = T))/(max(O1, na.rm = T) - min(O1, na.rm = T))*100
  ) %>%
  head(10)
```

::::
:::::
:::

## [Examples: Vector Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

-   Instead we could make this a function because I don't want to talk about how long it took me to do that copy-pasting 😭

::: fragment
::::::: columns
:::: {.column width="75%"}

```{r, eval = F}
pomp <- function(x, na) (x - min(x, na.rm = na))/(max(x, na.rm = na) - min(x, na.rm = na))*100

bfi %>%
  select(matches("1$")) %>%
  mutate(
    E1 = pomp(E1, T)
    , A1 =  pomp(A1, T)
    , C1 =  pomp(C1, T)
    , N1 =  pomp(N1, T)
    , O1 =  pomp(O1, T)
  ) %>%
  head(10)
```

::::
:::: {.column width="25%"}

```{r, echo = F}
pomp <- function(x, na) (x - min(x, na.rm = na))/(max(x, na.rm = na) - min(x, na.rm = na))*100
bfi %>%
  select(matches("1$")) %>%
  mutate(
    E1 = pomp(E1, T)
    , A1 =  pomp(A1, T)
    , C1 =  pomp(C1, T)
    , N1 =  pomp(N1, T)
    , O1 =  pomp(O1, T)
  ) %>%
  head(10)
```

::::
::::::
:::

## [Examples: Vector Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

-   And remember, we could simplify this even more with `mutate_at()`!
-   And I know in this example it saved us about four lines of code, but having a function like this that you can just reference, especially in conjunction with `mutate_at()` saves you ***time*** in the short and long run.

::: fragment
::::::: columns
:::: {.column width="75%"}

```{r, eval = F}
pomp <- function(x, na) (x - min(x, na.rm = na))/(max(x, na.rm = na) - min(x, na.rm = na))*100

bfi %>%
  select(matches("1$")) %>%
  mutate_at(vars(matches("1$")), pomp, na = T) %>%
  head(10)
```
::::

:::: {.column width="25%"}
```{r, echo = F}
pomp <- function(x, na) (x - min(x, na.rm = na))/(max(x, na.rm = na) - min(x, na.rm = na))*100

bfi %>%
  select(matches("1$")) %>%
  mutate_at(vars(matches("1$")), pomp, na = T) %>%
  head(10)
```
::::
::::::
:::

## [Writing Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   The first step in writing a function is to find the *pattern*:
-   Let's look back at the code above:

::: fragment
```{r, eval = F}
E1 = (E1 - min(E1, na.rm = T))/(max(E1, na.rm = T) - min(E1, na.rm = T))*100
A1 = (A1 - min(A1, na.rm = T))/(max(A1, na.rm = T) - min(A1, na.rm = T))*100
C1 = (C1 - min(C1, na.rm = T))/(max(C1, na.rm = T) - min(C1, na.rm = T))*100
N1 = (N1 - min(N1, na.rm = T))/(max(N1, na.rm = T) - min(N1, na.rm = T))*100
O1 = (O1 - min(O1, na.rm = T))/(max(O1, na.rm = T) - min(O1, na.rm = T))*100
```
:::

-   Do you see the pattern?

::: fragment
```{r, eval = F}
(█ - min(█, na.rm = TRUE)) / (max(█, na.rm = TRUE) - min(█, na.rm = TRUE))*100
```
:::

## [Writing Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

We need three things to turn this into a function:

```{r, eval = F}
(█ - min(█, na.rm = TRUE)) / (max(█, na.rm = TRUE) - min(█, na.rm = TRUE))*100
```

1.  **A name.** Here we’ll use pomp because this function rescales a vector to lie between POMP scores
2.  **The arguments.** The arguments are things that vary across calls and our analysis above tells us that we have just one. We’ll call it x because this is the conventional name for a numeric vector.
3.  **The body.** The body is the code that’s repeated across all the calls.

## [Writing Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   The template of a function looks something like this:

::: fragment
```{r, eval = F}
name <- function(arguments){
  body
}
```
:::

## [Writing Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Although for local functions, I suggest something like this:

::: fragment
```{r, eval = F}
name <- function(
    argument1 # description of first argument
    , argument2 # description of second argument
    ){
  body
}
```
:::

-   That's not going to fit well on my slides, so I will minimally do it in class.

## [Writing Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Following this, we return to our function:

::: fragment
```{r}
pomp <- function(x, na) {
  (x - min(x, na.rm = na))/(max(x, na.rm = na) - min(x, na.rm = na))*100
}
```
:::

## [Function Efficiency]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/orange-top.png" background-size="100%"}

-   For most of our purposes, we may not care that much about how fast our code is
-   Unless you're working with huge data, things typically run fast
-   But with a function like POMP, I may need to run it on 100+ variables each with 10,000+ observations, so speed may be a consideration
-   So how do you speed up your functions?
    1.  Avoid computing something twice
    2.  Try to anticipate where things could go wrong

## [Function Efficiency]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/orange-top.png" background-size="100%"}

-   So in our function, we calculate `min()` twice and `min()` once
-   We could just calculate `range()` instead, which reduces that computation time
-   Watch (note you have to run this in your console to see the difference)!

::::::: columns
:::: column
::: fragment
```{r}
pomp <- function(x) {
  (x - min(x, na.rm = T))/(max(x, na.rm = T) - min(x, na.rm = T))*100
}

start <- Sys.time()
tmp <- bfi %>%
  select(matches("1$")) %>%
  mutate_at(vars(matches("1$")), pomp)
(diff1 <- Sys.time() - start)
```
:::
::::

:::: column
::: fragment
```{r}
pomp <- function(x) {
  rng <- range(x, na.rm = T)
  (x - rng[1])/(rng[2] - rng[1])*100
}

start <- Sys.time()
tmp <- bfi %>%
  select(matches("1$")) %>%
  mutate_at(vars(matches("1$")), pomp)
(diff2 <- Sys.time() - start)
```
:::
::::
:::::::

## [Function Efficiency]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/orange-top.png" background-size="100%"}

-   But those differences are really minimal right? Just `r round(as.numeric(diff1 - diff2), 4)` seconds.
-   But across 150 variables that's `r round(as.numeric(diff1 - diff2)*150, 4)` seconds.

## [Cautionary note on outputs]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/orange-top.png" background-size="100%"}

-   The function we wrote above could be a "mutate()" function because the output is the same length / size as the input.
-   If you write a function that drops cases, you will get an error because R can't make the resulting vector fit back into the data frame
-   Or, if you write a function that returns length `1` within mutate, it will just repeat that value for every row
-   So, just be careful to think about what the output is and make sure that it matches the use case you are working with!

## [More notes on outputs]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/orange-top.png" background-size="100%"}


-   You can output anything you want from a function!

::::: columns
::: column
-   data frames
-   vectors
-   single values
-   strings
-   lists (very common for base R and package functions!!)
:::

::: column
-   model objects
-   plots
-   tables
-   etc.!
:::
:::::

## [You Try:]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/mint-top.png" background-size="100%"}

\vspace{2em}

Try turning the following into functions:

```{r, eval = F}
mean(is.na(x))
mean(is.na(y))
mean(is.na(z))

x / sum(x, na.rm = TRUE)
y / sum(y, na.rm = TRUE)
z / sum(z, na.rm = TRUE)

round(x / sum(x, na.rm = TRUE) * 100, 1)
round(y / sum(y, na.rm = TRUE) * 100, 1)
round(z / sum(z, na.rm = TRUE) * 100, 1)

(x - mean(x, na.rm = T))/sd(x, na.rm = T)
(y - mean(y, na.rm = T))/sd(y, na.rm = T)
(z - mean(z, na.rm = T))/sd(z, na.rm = T)
```

## [You Try (Solutions)]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/mint-top.png" background-size="100%"}

\vspace{2em}

```{r}
prop_missing <- function(x) mean(is.na(x))

prop_total   <- function(x) x / sum(x, na.rm = T)

round_prop   <- function(x) round(prop_total(x)*100, 1)

z_score      <- function(x) (x - mean(x, na.rm = T))/sd(x, na.rm = T)
```

## [Data Frame Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   All the functions we've covered so far are vector functions (i.e. they input a vector, not a matrix or data frame and work well within `mutate()` calls)
-   But if you have a long string of dplyr verbs, it can be useful to put these into a data frame function where you provide a data frame input and flexible way of naming the columns
-   I'm going to give a brief example here, but suggest that you check out more in [*R for Data Science*](https://r4ds.hadley.nz/functions#data-frame-functions).

## [Data Frame Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   Let's start with a brief example. To do it, let's first make the bfi data frame long

::: fragment
```{r}
bfi_long <- bfi %>%
  rownames_to_column("SID") %>%
  pivot_longer(
    cols = c(-SID, -education, -gender, -age)
    , names_to = c("trait", "item")
    , names_sep = -1
    , values_to = "value"
    , values_drop_na = T
  )
head(bfi_long, 8)
```
:::

## [Data Frame Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   So maybe I want to get means for each of the Big Five from this, so I write a function like this:

::: fragment
```{r, error=FALSE, eval=FALSE}
grouped_mean <- function(df, group_var, mean_var) {
  df %>%
    group_by(group_var) %>%
    summarize(mean(mean_var))
}

bfi_long %>% grouped_mean(trait, value)
```
:::

## [Data Frame Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

:::::: columns
::: {.column width="40%"}
-   That didn't work because we can't specify grouping variables like that
-   To get around it, we have to use something called **embracing**: We have to wrap the variables like this {{ trait }}
-   This just nudges the dplyr functions to look inside the data frame for the column you specify
:::

::: {.column width="60%"}
::: fragment
```{r}
tidy_describe <- function(df, var) {
  df %>%
    summarize(
      mean   = mean({{ var }},   na.rm = TRUE),
      sd     = sd({{ var }},     na.rm = TRUE),
      median = median({{ var }}, na.rm = TRUE),
      min    = min({{ var }},    na.rm = TRUE),
      max    = max({{ var }},    na.rm = TRUE),
      n      = n(),
      n_miss = sum(is.na({{ var }})),
      .groups = "drop"
      )
}

bfi_long %>% 
  group_by(trait) %>%
  tidy_describe(value)
```
:::
::::
::::::

## [Data Frame Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   Or remember when I had us getting counts and proportions for continuous x categorical relationships?

::: fragment
:::::: columns
:::: {.column width="60%"}
```{r, eval = F}
count_prop <- function(df, var, sort = FALSE) {
  df %>%
    count({{ var }}, sort = sort) %>%
    mutate(prop = n / sum(n))
}

bfi_long %>% 
  count_prop(education)
```
::::
:::: {.column width="40%"}
```{r, echo = F}
count_prop <- function(df, var, sort = FALSE) {
  df %>%
    count({{ var }}, sort = sort) %>%
    mutate(prop = n / sum(n))
}

bfi_long %>% 
  count_prop(education)
```
::::
::::::
:::

## [Wrap-Up]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   This is just a brief introduction to functions.
-   Functions are absolutely essential part of workflows, and you'll see them pop up in every lesson from here on out (and you already saw them pop up in previous lessons)
-   As we continue to see them, I'll ramp up their complexity, showing you how to write functions for estimating models, model predictions, figures, tables, and more

# Iteration and purrr {background-color="#00674F"}

## [Iteration]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

Iteration is everywhere. It underpins much of mathematics and statistics. If you've ever seen the $\Sigma$ symbol, then you've seen (and probably used) iteration.

Reasons for iteration:

- reading in multiple files from a directory\
- running the same operation multiple times\
- running different combinations of the same model\
- creating similar figures / tables / outputs

## [`for` loops]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

Enter `for` loops. `for` loops are the "OG" form of iteration in computer science. The basic syntax is below. Basically, we can use a for loop to loop through and print a series of things.

::: fragment
```{r basic loop}
for(i in letters[1:4]){
  print(i)
}
```
:::

-   The code above "loops" through 5 times, printing the iteration letter.

## [`_apply()` family]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/yellow-top.png" background-size="100%"}

-   A somewhat faster version of `for` loops comes from the `_apply()` family of functions, including:
    -   `apply()`, `lapply()`, `sapply()`, and `mapply()`. Unlike `for` loops, these are vectorized, which makes them more efficient.

::: fragment
```{r apply functions}
lapply(letters[1:4], print)
sapply(letters[1:4], print)
mapply(print, letters[1:4])
```
:::

## [`purrr` and `_map_()` functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

-   Today, though, we'll focus on the `map()` family of functions, which is the functions through which `purrr` iterates.

::: fragment
```{r map functions}
map(letters[1:4], print)
```
:::

-   **For a more thorough comparison of `for` loops, the `_apply()` family, and `_map_()` functions, see https://jennybc.github.io/purrr-tutorial/**

## [`purrr` and `_map_()` predicates]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

::::::: columns
:::: column
-   Today, though, we'll focus on the `purrr::map()` functions 

::: fragment
```{r map functions 2, eval = T}
map(letters[1:4], print)
```
:::

-   Note that this returns a list, which we may not always want.

::::

:::: column
-   With `purrr`, we can change the kind of output of `map()` by adding a predicate, like `lgl`, `dbl`, `chr`, and `df`.
-   So to print just the characters, we'd call `map_chr()`:

::: fragment
```{r basic map_chr}
map_chr(letters[1:4], print)
```
:::

-   Note that it also returns the vector and prints each letter iteratively.
::::
:::::::

## [`purrr` and `_map_()` antecedents]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

-   How many mappings?
    -   Single mapping: `map_()`\
    -   Parallel (2) mapping(s): `map2_()`\
    -   3 or more mappings: `pmap_()`

::: fragment
```{r}
map2_chr(letters[1:4], 1:4, paste)
```
:::

-   Note here that we can use `map2()` and `pmap()` with the predicates from above.

## [Use Cases]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

1.  Reading Data\
2.  Cleaning Data
3.  Running Models\
4.  (Plotting Figures - Week 8)\
5.  (Creating Tables - Week 8)

## [Reading Data]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

There are a number of different cases where `purrr` and `map()` maybe useful for reading in data including:

-   subject-level files for an experimental task\
-   subject- and task-level files for an experimental task
-   EMA data\
-   longitudinal data\
-   web scraping and text mining

## [Reading Data: Participant-Level EMA]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

For this first example, I'll show you how this would look with a `for` loop before I show you how it looks with `purrr`.

Assuming you have all the data in a single folder and the format is reasonably similar, you have the following basic syntax:

```{r simple reading loop, eval = F}
files <- list.files(data_path)
data <- list()
for(i in files){
  data[[i]] <- read_csv(i)
}
data <- combine(data)
```

## [Reading Data: Participant-Level EMA]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

-   This works fine in this simple case, but where `purrr` really shines in when you need to make modifications to your data before combining, whether this be recoding, removing missing cases, or renaming variables.

-   But first, the simple case of reading data. The code below will download a .zip file when you run it. Once, you do, navigate to your Desktop to unzip the folder. Open the R Project and you should be able to run the rest of the code

::: fragment
```{r get data}
data_source <- "https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/04-workshops/05-week5-purrr/week-5.zip"
data_dest <- "~/Desktop/week-5.zip"
download.file(data_source, data_dest)
```
:::

## [Reading Data: Participant-Level EMA]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

```{r read data ex1, eval = T}
df1 <- tibble(
  ID = list.files("data/example_1")
  ) %>%
  mutate(data = map(ID, ~read_csv(sprintf("data/example_1/%s", .)))) %>%
  unnest(data) 
```

-   The code above creates a list of ID's from the data path (files named for each person), reads the data in using the `map()` function from `purrr`, removes the ".csv" from the ID variable, then unnests the data, resulting in a data frame for each person.

## [List Columns and the Power of `purrr`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

-   On the previous slide, we saw a data frame **inside** of a data frame. This is called a list column within a nested data frame.

-   In this case, we created a list column using map, but one of the best things about `purrr` is how it combines with the `nest()` and `unnest()` functions from the `tidyr` package.

-   We'll return to `nest()` later to demonstrate how anything you would iterate across is also something we can `nest()` by in long format data frames.

## [Exercise: Reading in Participant-Level Data]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/mint-top.png" background-size="100%"}

\vspace{2em}

-   Now, we're going to combine with what we learned about last time with codebooks.

::: fragment
```{r codebook}
codebook <- read_csv("data/codebook_ex1.csv")
codebook
```
:::

## [Exercise: What next?]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/mint-top.png" background-size="100%"}

\vspace{2em}

:::::: columns
:::: column
-   Now, that we have a codebook, what are the next steps?

::: fragment
```{r read data ex1 2, eval = T}
df1 <- tibble(
  ID = list.files("data/example_1")
  ) %>%
  mutate(data = map(ID, ~read_csv(sprintf("data/example_1/%s", .)))) %>%
  unnest(data) 
print(df1, n = 6, width = 50)
```
:::
::::

::: column
1.  pull old names in raw data from codebook\
2.  pull new names from codebook\
3.  select columns from codebook in loaded data\
4.  rename columns with new names
:::
::::::

## [Exercise: Reading in Participant-Level Data]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/mint-top.png" background-size="100%"}

::: fragment
```{r read data complex ex1, eval = T}
#| code-line-numbers: "|1-2|3-5|6-9|10|1-2,11"
old.names <- codebook$old_name # pull old names in raw data from codebook  
names(old.names) <- codebook$item_name # pull new names from codebook  
df1 <- tibble(
  ID = list.files("data/example_1")
  ) %>%
  mutate(
    data = map(ID, ~read_csv(sprintf("data/example_1/%s", .)))
    , ID = str_remove_all(ID, ".csv")
    ) %>%
  unnest(data) %>%
  select(ID, count, all_of(old.names)) # select & rename columns from codebook in loaded data  
head(df1, 6)
```
:::

## [Descriptives: Within-Person Correlations]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

With these kinds of data, the first thing, we may want to do is look at within-person correlations, which we can do with `purrr`.

```{r}
#| code-line-numbers: "|1|3-6|7"
cor_fun <- function(x) cor(x %>% select(-count), use = "pairwise")

nested_r <- df1 %>%
  group_by(ID) %>%
  nest() %>%
  ungroup() %>%
  mutate(r = map(data, cor_fun))
head(nested_r, 6)
```

## [Descriptives: Within-Person Correlations]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

We can access it like a list:

```{r}
nested_r$data[[1]]
```

## [Descriptives: Within-Person Correlations]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   But we can't easily (well, nicely) just `unnest()` a matrix
-   We'll lose a lot of information (e.g., column and row names) along the way
-   So what do we do?

::: fragment
```{r}
nested_r %>%
  select(-data) %>%
  unnest(r) %>%
  print(n=10)
```
:::

## [Descriptives: Within-Person Correlations]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

::::::: columns
:::: column
-   Write a function, of course!

::: fragment
```{r}
#| code-line-numbers: "1-10|2-5|6-9|12-13"
cor_fun <- function(x) {
  r <- cor(
    x %>% select(-count)
    , use = "pairwise"
    )
  r %>% 
    data.frame() %>%
    rownames_to_column("var") %>%
    as_tibble()
}

nested_r <- nested_r %>%
  mutate(r = map(data, cor_fun))
```
:::
::::

:::: column
::: fragment
```{r, echo = F}
head(nested_r, 10)
```
:::
::::
:::::::

## [Descriptives: Within-Person Correlations]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   Let's try unnesting again:

::: fragment
```{r}
#| code-line-numbers: "|1-2|3|5"
nested_r %>%
  select(-data) %>%
  unnest(r) %>% 
  arrange(desc(var))
```
:::

## [Descriptives: Within-Person Correlations]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   There's more I would usually do here to format a correlation table for each participant and output the file as a PDF or html so I can post it on GitHub / OSF
-   But we'll get there in Week 8/9!

## [Descriptives: Means, sds, etc.]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

::::::: columns
:::: column
::: fragment
```{r}
tidy_describe <- function(df) {
  df %>%
    pivot_longer(
      -count
      , names_to = "item"
      , values_to = "value"
      , values_drop_na = T
      ) %>%
    group_by(item) %>%
    summarize(
      mean   = mean(value,   na.rm = TRUE),
      sd     = sd(value,     na.rm = TRUE),
      median = median(value, na.rm = TRUE),
      min    = min(value,    na.rm = TRUE),
      max    = max(value,    na.rm = TRUE),
      n      = n(),
      n_miss = sum(is.na(value)),
      .groups = "drop"
      )
}
```
:::
::::

:::: column
::: fragment
```{r}
nested_r <- nested_r %>%
  mutate(desc = map(data, tidy_describe)) 
nested_r %>% print(n = 5, width = 50)
```
:::
::::
:::::::

## [Descriptives: Means, sds, etc.]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

```{r}
nested_r %>%
  select(-data, -r) %>%
  unnest(desc)
```

## [Models]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

-   We can put essentially anything into a nested data frame.
-   The magic happens because everything is indexed by the other columns in the data frame, so we can keep track of it
-   And unlike a normal list, we aren't stuck with nested list structures that are really hard to parse and navigate through
-   Next, I'm going to show you how to use purrr with models
-   Modeling is not a focus of this class, but I want to demonstrate this as a \~workflow\~ because it completely revolutionized mine!

## [Models]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

::::::: columns
:::: column
-   But first, we need to format our data:

::: fragment
```{r}
df_long <- df1 %>%
  select(-satisfaction) %>%
  pivot_longer(
    cols = c(-count, -ID)
    , names_to = c("trait", "item")
    , names_sep = "_"
    , values_to = "value"
    , values_drop_na = T
    )
```
:::
::::

:::: column
::: fragment
```{r}
df_long
```
:::
::::
:::::::

## [Models]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

:::::: columns
::: column
To create composites, we'll:\

1. separate traits from items 
2. `group_by()` trait, count, and ID 
3. calculate the composites using `summarize()`

:::

:::: column
::: fragment
```{r}
df_long <- df_long %>%
  group_by(ID, count, trait) %>%
  summarize(value = mean(value)) %>%
  ungroup() %>%
  left_join(df1 %>% select(ID, count, satisfaction))
df_long
```
:::
::::
::::::

## [Models]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

Then we'll get within-person centered values using our own little function!

::: fragment
:::::: columns
:::: column 

```{r, eval = F}
#| code-line-numbers: "|1|3-4|5|6"
center <- function(x) x - mean(x, na.rm = T)

df_long <- df_long %>%
  group_by(ID, trait) %>%
  mutate(value_c = center(value)) %>%
  ungroup()
df_long
```
::::
:::: column 

```{r, echo = F}
center <- function(x) x - mean(x, na.rm = T)

df_long <- df_long %>%
  group_by(ID, trait) %>%
  mutate(value_c = center(value)) %>%
  ungroup()
df_long
```
::::
::::::
:::

## [Models]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

And grand-mean centered within-person averages

::: fragment
:::::: columns
:::: column 
```{r, eval = F}
#| code-line-numbers: "|2-3|4-5|6"
df_long <- df_long %>%
  group_by(ID, trait) %>%
  mutate(value_gmc = mean(value)) %>%
  group_by(trait) %>%
  mutate(value_gmc = center(value_gmc)) %>%
  ungroup()
df_long
```
::::
:::: column 
```{r, echo = F}
df_long <- df_long %>%
  group_by(ID, trait) %>%
  mutate(value_gmc = mean(value)) %>%
  group_by(trait) %>%
  mutate(value_gmc = center(value_gmc)) %>%
  ungroup()
df_long %>% print(width = 50)
```
::::
:::::: 
:::

## [Models]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

And now we are ready to run our models. But first, we'll `nest()` our data.

::: fragment
```{r}
#| code-line-numbers: "|1|2|3|4"
nested_mods <- df_long %>%
  group_by(trait) %>%
  nest() %>%
  ungroup() 
nested_mods
```
:::

## [Models]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

And now run the models.

::: fragment
```{r}
#| code-line-numbers: "|1|7"
run_model <- function(d) lmer(satisfaction ~ value_c * value_gmc + (1 | ID), data = d)

nested_mods <- df_long %>%
  group_by(trait) %>%
  nest() %>%
  ungroup() %>% 
  mutate(model = map(data, run_model))
nested_mods
```
:::

## [Models]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

And get data frames of the results:

::: fragment
```{r}
#| code-line-numbers: "|3-11|4|5-7|5|6|7|8|1|9|10|13-14"
sprintfna <- function(x) ifelse(is.na(x), NA_character_, sprintf("%.2f", x))

tidy_tab <- function(m){
  tidy(m, conf.int = T) %>%
    mutate(pval = pnorm(abs(estimate/`std.error`), lower.tail = FALSE),
           p = round(pval, digits = 3),
           p = ifelse(pval < .001, "p &lt; .001", paste0("p = ", p))) %>%
    mutate_at(vars(estimate, conf.low, conf.high), sprintfna) %>%
    mutate(CI = ifelse(is.na(conf.low), "", sprintf("[%s,%s]", conf.low, conf.high))) %>%
    dplyr::select(term, estimate, CI, p)
}

nested_mods <- nested_mods %>%
  mutate(tidy = map(model, tidy_tab))
nested_mods
```
:::

## [Models: Unnesting]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

Which we can print into pretty data frames

::: fragment
```{r}
nested_mods %>%
  select(trait, tidy) %>%
  unnest(tidy)
```
:::

## [Models: Unnesting & Tabling]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

Which we can pretty easily turn into tables:

::: fragment
```{r}
mod_tab <- nested_mods %>%
  select(trait, tidy) %>%
  unnest(tidy) %>%
  pivot_wider(
    names_from = "trait"
    , names_glue = "{trait}_{.value}"
    , values_from = c("estimate", "CI", "p")
  ) %>%
  select(term, starts_with("E"), starts_with("A"), starts_with("C"), starts_with("N"), starts_with("O"))

mod_tab
```
:::

## [Models: Unnesting & Tabling]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

```{r}
hdr <- c(1, rep(3, 5))
names(hdr) <- c(" ", "Extraversion", "Agreeableness", "Conscientiousness", "Neuroticism", "Openness")

tab <- mod_tab %>%
  kable(.
        , "html"
        , escape = F
        , col.names = c("Term", rep(c("<em>b</em>", "CI", "<em>p</em>"), times = 5))
        , align = c("r", rep("c", 15))
        , caption = "<strong>Table 1</strong><br><em>Multilevel Model Estimates of Between- and Within-Person Big Five-State Satisfaction Associations"
        ) %>%
  kable_classic(full_width = F, html_font = "Times", font_size = 15) %>%
  add_header_above(hdr)
```

## [Models: Unnesting & Tabling]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

```{r}
tab
```

## [Models: Unnesting & Tabling]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

```{r}
hdr <- c(1, rep(3, 5))
names(hdr) <- c(" ", "Extraversion", "Agreeableness", "Conscientiousness", "Neuroticism", "Openness")
terms <- tibble(
  old = c("(Intercept)", "value_c", "value_gmc", "value_c:value_gmc", "sd__(Intercept)", "sd__Observation")
  , new = c("Intercept", "Within", "Between", "Cross-level", "Random Intercept", "Random Slope")
  )

tab <- mod_tab %>%
  mutate(term = factor(term, levels = terms$old, labels = terms$new)) %>%
  kable(.
        , "html"
        , escape = F
        , col.names = c("Term", rep(c("<em>b</em>", "CI", "<em>p</em>"), times = 5))
        , align = c("r", rep("c", 15))
        , caption = "<strong>Table 1</strong><br><em>Multilevel Model Estimates of Between- and Within-Person Big Five-State Satisfaction Associations"
        ) %>%
  kable_classic(full_width = F, html_font = "Times", font_size = 15) %>%
  add_header_above(hdr)
```

## [Models: Unnesting & Tabling]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

```{r}
tab
```

