---
title: "Problem Set #4"
author: "INSERT YOUR NAME HERE"
date: "insert date here"
urlcolor: blue
format: 
  html:
    code-tools: true
    code-copy: true
    code-line-numbers: true
    code-link: true
    theme: united
    highlight-style: tango
    df-print: paged
    code-fold: show
    toc: true
    toc-float: true
    self-contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", highlight = TRUE)
```

```{r}
library(psych)
library(lme4)
library(tidyverse)

df <- psych::bfi %>%
  rownames_to_column("id")
```


# Part 1: Using Functions

Functions are really useful when you have to do something many times. 

## Practicing Function Writing 

Write a function to z-standardize data ((observed - mean)/standard deviation). Apply it to at least 2 variables in your data. (Note: if your data are in long form, you will either need to make your data wide by some key variable (stimuli, question, etc. or will need to use your function on a grouped data frame))

```{r z-scoring}
# your code here
z_score <- function(x) (x - mean(x, na.rm = T))/sd(x, na.rm = T)
df <- df %>%
  mutate_at(vars(matches("[AECNO]\\d")), lst(z = z_score))
```

Now look at your descriptives, do the descriptives suggest your function worked correctly? (Hint: if in long form, make sure you're looking at the descriptives grouped)

```{r z-desc}
# your code here
df %>%
  select(id, matches("[AECNO]\\d")) %>%
  describe()
```

# Part 2: Iteration

In class, we talked about iteration as `for` loops, `lapply()`, and `purrr:::map()`. But we've actually been doing iteration for weeks using functions like `mutate_at()` and `mutate_all()`. Another alternative is using `mutate(across())`, which works similar to `mutate_at()` but is more generic. So for example the code below reverse scores the BFI items that are negatively keyed: 

```{r, eval = F}
psych::bfi %>%
  mutate(across(c(A1, C4, C5, E1, E2, O2, O5), ~6 - .))
```

1. Using at least two methods of iteration, apply a function to multiple columns, subsets (e.g., participants, stimuli, waves, etc.). For example, as above use two methods to mutate multiple columns (hint: see `?apply` or `?map_dbl`) or to calculate descriptives, correlations, etc. (hint: see `?lappy` and `?map). 

```{r method 1}
# your code here
apply(df %>% select(matches("E\\d$")), 2, z_score)
```

```{r method 2}
# your code here
map_df(df %>% select(matches("E\\d$")), z_score)
```

Do you get the same results? If not, why? 

**Yes, but the output are different object classes (array v. data frame)**

2. Write a function that estimates multiple descriptives (mean, median, sd, min, max, n, n missing). Using any form of iteration, apply that function to all continuous variables in your data frame. (Hints: you could (1) pivot your data to long and `group_by()` item, either nesting and applying your function or writing a data frame function or (2) use a function like apply() or across to estimate them. Note the format challenges you experience [e.g., errors, your data are super wide].)

Ultimately, you want to end up with a data frame with items / indicators, etc. as rows (indexed by a column) and columns for each of the descriptives. 

# Part 3: Strings

## Working with strings

1. Load the following packages in the code chunk below: `tidyverse` and `lubridate`.

```{r, message=F}
desc_fun <- function(x){
  x %>% 
    summarize(
      N = n()
      , n_missing = sum(is.na(value))
      , mean = mean(value, na.rm = T)
      , median = median(value, na.rm = T)
      , sd = sd(value, na.rm = T)
      , min = min(value, na.rm = T)
      , max = max(value, na.rm = T)
    )
}

df %>% 
  select(id, ends_with("z")) %>%
  pivot_longer(
    cols = -id
    , names_to = "item"
    , values_to = "value"
  ) %>%
  group_by(item) %>%
  nest() %>%
  ungroup() %>%
  mutate(desc = map(data, desc_fun)) %>%
  select(-data) %>%
  unnest(desc)
```


2. Using `str_c()` and the following objects as input, create the string: `"Roses are red, Violets are blue"`  

    - We encourage you to first sketch out what you want to do on some scratch paper.
    - Recall from the lecture example on "Using str_c() on vectors of different lengths", when multiple vectors of different length are provided in the str_c() function, the elements of shorter vectors are recycled. See below.
    
```{r eval=FALSE}
str_c("@", c("emorie ", "sgtpepper ", "apple "), sep = "", collapse = ",")

#[1] "@emorie ,@sgtpepper ,@apple "
```
    
    - Now try it yourself.
```{r}
vec_1 <- c("Roses", "Violets")
vec_2 <- c("red", "blue")

str_1 <- "are"

# Write your code here
str_c(vec_1, str_1, vec_2, sep = " ", collapse = ", ")

```


3. [Pig Latin](https://en.wikipedia.org/wiki/Pig_Latin) is a language game in which the first consonant of each word is moved to the end of the word, then `"ay"` is appended to create a suffix. For example, the word `"Wikipedia"` would become `"Ikipediaway"`.

    - Using `str_c()` and `str_sub()`, turn the given `pig_latin` vector into the string: `"igpay atinlay"`
    - We encourage you to first sketch out what you want to do on some scratch paper.
        - First, think about what the final outcome will look like.
        - Then, think about how you can get there. Play around with the `str_sub()` function. What happens when you include different values in the `str_sub()` function?  
    - this is low-key the trickiest question in the problem set. So if you get stuck, ask a question to your group or github and move on. and come back to it later.

```{r}
pig_latin <- c('pig', 'latin')

# Write your code here
str_c(
  str_sub(pig_latin, start = 2, end = -1)
  , str_sub(pig_latin, start = 1, end = 1) 
  , "ay"
)


```


4. Using `str_c()` and `str_sub()`, decode the given `secret_message`. Your output should be a string.  

    - Follow the same logic from above.
    - Sketch out what you want to do on some scratch paper. Break it down step by step. Play around with different values for the `str_sub()` function. 
    
```{r}
secret_message <- c('ollowfay', 'ouryay', 'earthay')

# Write your code here
str_c(
  str_sub(secret_message, start = -3, end = -3)
  , str_sub(secret_message, end = -4)
)
```


## Working with Twitter data  


1. You will be using Twitter data we fetched from the following Twitter handles: `UniNoticias`, `FoxNews`, and `CNN`.  

    - This data has been saved as an Rdata file.
    - Use the `load()` and `url()` functions to download the `news_df` dataframe from the url: `https://github.com/emoriebeck/psc203a-data-FQ26/raw/main/05-assignments/04-ps4/twitter_news.RData`
    - Report the dimensions of the `news_df` data frame (rows and columns). Use the `dim()` function.  


```{r}
# Write your code here
load(url("https://github.com/emoriebeck/psc203a-data-WQ26/raw/refs/heads/main/05-assignments/04-ps4/twitter_news.RData"))
dim(news_df)
```


2. Subset your dataframe `news_df` and create a new dataframe called `news_df2` keeping only the following variables: `user_id`, `status_id`, `created_at`, `screen_name`, `text`, `followers_count`, `profile_expanded_url`.  

    - Note in the following questions we will ask you to create a new column and that means you have to assign `<-` the new changes you are making to the existing dataframe `news_df2`. Ex. `news_df2 <- news_df %>% mutate(newvar = mean(oldvar))`

```{r}
cls <- c("user_id", "status_id", "created_at", "screen_name", "text", "followers_count", "profile_expanded_url")
# Write your code here
news_df2 <- news_df %>%
  select(cls)
```


3. Create a new column in `news_df2` called `text_len` that contains the length of the character variable `text`.

    - What is the `class` and `type` of this new column? Make sure to include your code in the code chunk below.
      - __ANSWER__: 
    
```{r}
# Write your code here
news_df2 <- news_df2 %>%
  mutate(text_len = str_length(text))
class(news_df2$text_len)
typeof(news_df2$text_len)
```


4. Create an additional column in `news_df2` called `handle_followers` that stores the twitter handle and the number of followers associated with that twitter handle in a string. For example, the entries in the `handle_followers` column should look like this: `@[twitter_handle] has [number] followers.`

    - What is the `class` and `type` of this new column? Make sure to include your code in the code chunk below.
      - __ANSWER__: 

```{r}
# Write your code here
news_df2 <- news_df2 %>%
  mutate(handle_followers = sprintf("@%s has %i followers", screen_name, followers_count))
class(news_df2$handle_followers)
typeof(news_df2$handle_followers)
```


5. Lastly, create a column in `news_df2` called `short_web` that contains a short version of the `profile_expanded_url` without the `http://www.` part of the url. For example, the entries in that column should look something like this: `nytimes.com`.

```{r}
# Write your code here
news_df2 <- news_df2 %>%
  mutate(short_web = str_remove_all(profile_expanded_url, "http://www."))
```

# Part 4: Dates 
## Working with dates/times

1. Using the column `created_at`, create a new column in `news_df2` called `dt_chr` that is a character version of `created_at`.

    - What is the `class` of the `created_at` and `dt_chr` columns? Make sure to include your code in the code chunk below.
      - __ANSWER__: Character

```{r}
# Write your code here
news_df2 <- news_df2 %>%
  mutate(dt_chr = as.character(created_at))
class(news_df2$dt_chr)
typeof(news_df2$dt_chr)
```


2. Create another column in `news_df2` called `dt_len` that stores the length of `dt_chr`.

```{r}
# Write your code here
news_df2 <- news_df2 %>%
  mutate(dt_len = as.character(dt_chr))
news_df2$dt_chr
```


3. Next, create additional columns in `news_df2` for each of the following date/time components:

    a. Create a new column `date_chr` for date (e.g. `2020-03-26`) using the column `dt_chr` and the `str_sub()` function.
    a. Do the same for year `yr_chr` (e.g. `2020`).
    a. Do the same for month `mth_chr` (e.g. `03`).
    a. Do the same for day `day_chr` (e.g. `26`).
    a. Do the same for time `time_chr` (e.g. `22:41:09`).  

```{r}
# Write your code here
news_df2 <- news_df2 %>%
  mutate(
    date_chr = str_sub(dt_chr, start = 1, end = 10)
    , yr_chr = str_sub(dt_chr, start = 1, end = 4)
    , mth_chr = str_sub(dt_chr, start = 6, end = 7)
    , mthday_chr_chr = str_sub(dt_chr, start = 9, end = 10)
    , time_chr = str_sub(dt_chr, start = 12, end = 19)
    )
```


4. Using the column we created in the previous question `time_chr`, create additional columns in `news_df2` for the following time components: 

    a. Create a new column `hr_chr` for hour (e.g. `22`) using the column `time_chr` and the `str_sub()` function.
    a. Do the same for minutes `min_chr` (e.g. `41`).
    a. Do the same for seconds `sec_chr` (e.g. `09`).

```{r}
# Write your code here
news_df2 <- news_df2 %>%
  mutate(
    hr_chr = str_sub(time_chr, start = 1, end = 2)
    , min_chr = str_sub(time_chr, start = 4, end = 5)
    , sec_chr = str_sub(time_chr, start = 7, end = 8)
  )
```



5. Now let's get some practice with the `lubridate` package.  

    a. Using the `year()` function from the `lubridate` package, create a new column in `news_df2` called `yr_num` that contains the year (e.g. `2020`) extracted from `date_chr`.
    a. Do the same for month `mth_num`.
    a. Do the same for day `day_num`.
    a. Do the same for hour `hr_num`, but extract from `created_at` column instead of `date_chr`.
    a. Do the same for minutes `min_num`. 
    a. Do the same for seconds `sec_num`.

```{r}
# Write your code here
news_df2 <- news_df2 %>%
  mutate(
    yr_num = year(date_chr)
    , mth_num = month(date_chr)
    , day_num = day(date_chr)
    , hr_num = hour(date_chr)
    , min_num = minute(date_chr)
    , sec_num = second(date_chr)
  )
```

6. Using the **new numeric columns** (e.g. day_num, mth_num) you've created in the previous step, reconstruct the date and datetime columns. Namely, add the following columns to `news_df2`:

    a. Use `make_date()` to create new column called `my_date` that contains the date (year, month, day).
    a. Use `make_datetime()` to create new column called `my_datetime` that contains the datetime (year, month, day, hour, minutes, seconds).
    
    - What is the `class` of your `my_date` and `my_datetime` columns? Make sure to include your code in the code chunk below.
      - __ANSWER__: 

```{r}
# Write your code here
news_df2 %>%
  mutate(
    my_date = make_date(
      year = yr_num
      , month = mth_num
      , day = day_num
      )
    , my_datetime = make_datetime(
      year = yr_num
      , month = mth_num
      , day = day_num
      , hour = hr_num
      , min = min_num
      , sec = sec_num
      )
    )
```

The purpose of this problem set is for you to understand how the backslash escape character (`\`) works in R strings, as well as to practice writing regular expressions. You will be using the `str_view_all()` function to see all the matches from your regex. You'll get practice combining character classes, quantifiers, anchors, ranges, groups, and more to build your regular expressions for each question.

# Part 5: Regex
## Backslash (`\`) escape character

In this section, you will practice working with strings that include backslashes, such as for escaping characters or for writing special characters. You will be using both the `print()` and `writeLines()` functions to print out your string and compare the difference. This section is not specific to/does not involve regular expressions.

\textcolor{red}{\textbf{/1}}

1. Create a short string (could be a phrase or sentence) that contains both the single quote (`'`) and double quote (`"`) inside your string, and save it as an object called `string_with_quotes`. Use both `print()` and `writeLines()` to print out your string.  

    _Hint_: You will need to use a backslash to escape either the single quote (`'`) or the double quote (`"`) depending on if you used single or double quotes to enclose your string. 

```{r}
# Write your code here
string_with_quotes <- "This is the day's \"test\""
print(string_with_quotes)
writeLines(string_with_quotes)
```

2. Create a short string (could be a phrase or sentence) that contains both the tab and newline special characters, and save it to `string_with_spchars`. Use both `print()` and `writeLines()` to print out your string.

```{r}
string_with_spchars <- "This\tis just\nmean"
print(string_with_spchars)
writeLines(string_with_spchars)
# Write your code here
```

3. Create a string that contains your first name where each letter is separated by a backslash (e.g., `y\o\u\r\n\a\m\e`), and save it to `string_with_backslashes`. Use both `print()` and `writeLines()` to print out your string.

    _Hint_: Your `writeLines()` output should show single backslashes between each letter of your name.

```{r}
string_with_backslashes <- "e\\m\\o\\r\\i\\e"
print(string_with_backslashes)
writeLines(string_with_backslashes)
# Write your code here
```

4. With respect to the previous questions, explain in general why the output created by the `print()` function differs from the output created by the `writeLines()` function.

**One is parsing the syntax, while the other is not. **

## Matching characters

In this section and the next, you will practice writing regular expressions to match specific text. Use `str_view_all()` for all the following questions to show the matches.

1. Show all matches to single quotes (`'`) in your `string_with_quotes` that you created in Part I.

```{r}
str_view(string_with_quotes, pattern = "'")
# Write your code here
```

2. Show all matches to double quotes (`"`) in `string_with_quotes`.

```{r}
# Write your code here
str_view(string_with_quotes, '"')
```

3. Show all matches to tab characters in `string_with_spchars`.

```{r}
# Write your code here
str_view(string_with_spchars, "\\t")
```

4. Show all matches to newline characters in `string_with_spchars`.

```{r}
# Write your code here
str_view(string_with_spchars, "\\n")
```

5. Show all matches to backslashes (`\`) in `string_with_backslashes`.

```{r}
str_view(string_with_backslashes, "\\\\")
```


## Regular expressions

\textcolor{red}{\textbf{/1}}

1. Copy the following code to create the character vector `text`:

```{r}
    text <- c("In 5... 4... 3... 2...",
              "It can cost anywhere between $50 to $100 (... or even $1k!)",
              "These are parenthesis (), while these are brackets []... I think.")
```

```{r}
# Write your code here
```

2. Show all matches to a capital `I` at the beginning of the string.

```{r}
# Write your code here
str_view(text, pattern = "^I")
```

3. Show all matches to a period at the end of the string. 

```{r}
# Write your code here
str_view(text, pattern = "[.]$")
```

4. Show all matches to 1 or more digits.

```{r}
# Write your code here
str_view(text, pattern = "\\d+")
```

5. Show all matches to all dollar amounts, including the dollar sign and `k` if there is one (i.e., `$50`, `$100`, `$1k`)

```{r}
# Write your code here
str_view(text, pattern = "\\$(\\d+k|\\d+)")
```

6. Show all matches to ellipses (`...`)

```{r}
# Write your code here
str_view(text, pattern = "[.]{3}")
```

7. Show all matches to parentheses, including the contents between the parentheses if there are any.

```{r}
# Write your code here
str_view(text, pattern = "(?<=\\().*(?=\\))")
```

8. Show all matches to words (define words as containing only letters, upper or lowercase)

```{r}
# Write your code here
str_view(text, pattern = "[A-Za-z]+")
```

9. Show all matches to either a word that's 4 or more letters long _or_ ellipses.

```{r}
# Write your code here
str_view(text, pattern = "[A-Za-z]{4,}|\\.\\.\\.")
```

10. Show all matches to any digit or vowel (upper or lowercase) that repeats 2 times in a row (i.e., the same digit or vowel repeated twice in a row)

```{r}
# Write your code here
str_view(text, pattern = "([aeiouAEIOU0-9])\\1")
```

# Render to html and submit problem set  

**Render to html** by clicking the "Render" button near the top of your RStudio window (icon with blue arrow) 

-   Go to the Canvas --\> Assignments --\> Problem Set 4
-   Submit both .qmd and .html files\
-   Use this naming convention "lastname_firstname_ps#" for your .qmd and html files (e.g. beck_emorie_ps4.qmd & beck_emorie_ps4.html)

