---
title: "Week 6 - Review"
author: "Emorie D Beck"
format: 
  revealjs:
    incremental: true
    code-tools: true
    code-copy: true
    code-line-numbers: true
    code-link: true
    preview-links: true
    slide-number: true
    self-contained: true
    fig-height: 4
    fig-width: 6
    fig-align: center
    css: custom.css
    theme: psc290-23
    # highlight-style: atom-one-dark
    margin-left: "0"
    margin-right: "0"
    width: 1400
    # height: 900
    footer: "PSC 203a - Data Management and Cleaning"
    logo: "https://github.com/emoriebeck/psc290-data-viz-2022/raw/main/01-week1-intro/02-code/02-images/ucdavis_logo_blue.png"
editor: visual
editor_options: 
  chunk_output_type: console
---

```{r, echo = F}
library(knitr)
library(psych)
library(lme4)
library(broom)
library(broom.mixed)
library(kableExtra)
library(plyr)
library(tidyverse)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE,	
                      warning = FALSE,
                      results = 'show',
                      fig.width = 4, 
                      fig.height = 4, 
                      fig.retina = 3)
options(htmltools.dir.version = FALSE
        , knitr.kable.NA = "")
```

# Outline

2.  `dplyr`
3.  `tidyr`
4.  Functions
5.  `purrr`

# tidyverse: chaining (`%>%` or `|>`)  {background-color="black"}

# dplyr: select() and filter() {background-color="blue"}

## [1. `select()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

### Add or remove using `select()` helper functions.

:::::: columns
::: {.column width="40%"}
-   `starts_with()`\
-   `ends_with()`
-   `contains()`
-   `matches()`
-   `num_range()`
-   `one_of()`
-   `all_of()`
:::

:::: {.column width="60%"}
::: fragment
```{r, echo = T}
bfi |>
  select(starts_with("C"))
```
:::
::::
::::::

## [2. `filter()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%" .smaller}

-   Often times, when conducting research (experiments or otherwise), there are observations (people, specific trials, etc.) that you don't want to include.
-   We can use `filter()` with logical statements to include only rows that match certain conditions
-   We can refer to both bare quoted columns and objects in the global environment

::::: columns
::: column
-   `==` or `!=`
-   `<` or `<=`
-   `>` or `>=`
-   `%in%`
:::

::: column
-   `all_of()`
-   `one_of()`
-   `!`
-   `|` and `&`
:::
:::::

# tidyr: pivot_longer() and pivot_wider() {background-color="#00674F"}

## [1. `pivot_longer()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%" .smaller}

-   (Formerly `gather()`) Makes wide data long, based on a key <font size="5">
-   Core arguments:
    -   `data`: the data, blank if piped
    -   `cols`: columns to be made long, selected via `select()` calls
    -   `names_to`: name(s) of key column(s) in new long data frame (string or string vector)
    -   `values_to`: name of values in new long data frame (string)
    -   `names_sep`: separator in column headers, if multiple keys
    -   `values_drop_na`: drop missing cells (similar to `na.rm = T`) </font>

## [1. `pivot_longer()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%" .smaller}

### Why would I make my data longer?

-   Main reason: Columns names sometimes contain data.
-   Example: Billboard data has *time* information in column names

::: fragment
```{r}
data(billboard)
str(billboard)
```
:::

## [1. `pivot_longer()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%" .smaller}

### Why would I make my data longer?

```{r}
billboard |>
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    names_prefix = "wk",
    names_transform = as.numeric,
    values_to = "rank"
  )
```

## [1. `pivot_longer()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%" .smaller}

### Why would I make my data longer?

-   This doesn't just apply to longitudinal data. This is also important when thinking about iteration.
-   For example, if you variables can be grouped into different categories (covariates, IVs/predictors, DVs/outcomes, moderators, etc.), then your column names contain *implicit* data
-   The data below contain *both* time, variable, and category information

:::::: fragment
::::: columns
::: column
```{r}
# load the codebook
codebook <- read_csv("week6-codebook.csv") |>
    mutate(old_name = str_to_lower(old_name))
codebook %>% print(n = 6, width = 70)
```
:::

::: column
```{r}
old.names <- codebook$old_name # get old column names
names(old.names) <- codebook$new_name # get new column names
soep <- read_csv("week6-data.csv") |>
    select(all_of(old.names))
soep %>% print(n = 6, width = 70)
```
:::
:::::
::::::

## [1. `pivot_longer()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%" .smaller}

### Why would I make my data longer?

-   By pivoting our data longer, we can more easily extract information from the column names
-   Pivot the data below longer. Some hints:
    -   don't make the procedural or demographic variables long
    -   split the column names you make long into four chunks:
        -   "category", "label", "item_name", "year"
    -   drop na values

## [1. `pivot_longer()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%" .smaller}

### Why would I make my data longer?

By pivoting our data longer, we can more easily extract information from the column names

::: fragment
```{r}
soep_long <- soep |>
  pivot_longer(
    cols = c(-starts_with("Proc"), -starts_with("Dem"))
    , names_to = c("category",	"label",	"item_name",	"year")
    , names_pattern = "(.*)_(.*)_(.*)_(.*)"
    , values_to = "value"
    , values_drop_na = T
  ) |> mutate(year = as.numeric(year))
soep_long
```
:::

## [1. `pivot_longer()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%" .smaller}

### Why would I make my data longer?

-   Long format data are easier to clean (we'll come back to this but we'll create the cleaned data frame to use for merging practice)

::: fragment
```{r}
soep_big5 <- soep_long |>
  filter(category == "big5") |>
  mutate(value = mapvalues(value, seq(-8,0), rep(NA, 9))) |>
  drop_na(value) |>
  group_by(Proc_SID, label, year) |>
  summarize(value = mean(value)) |>
  ungroup()

soep_le <- soep_long |>
  filter(category == "le") |>
  mutate(value = mapvalues(value, seq(-8,1), c(rep(NA, 6), 0, NA, NA, 1))) |>
  drop_na(value) |>
  group_by(Proc_SID, label) |>
  summarize(value = sum(value)) |>
  ungroup()

soep_clean <- soep_big5 |>
  rename(trait = label, p_value = value) |>
  inner_join(
    soep_le |>
      rename(le = label, le_value = value)
  )
```
:::

## [2. `pivot_wider()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%" .smaller}

-   (Formerly `spread()`) Makes wide data long, based on a key <font size="6">
-   Core arguments:
    -   `data`: the data, blank if piped
    -   `names_from`: name(s) of key column(s) in new long data frame (string or string vector)
    -   `names_sep`: separator in column headers, if multiple keys
    -   `names_glue`: specify multiple or custom separators of multiple keys
    -   `values_from`: name of values in new long data frame (string)
    -   `values_fn`: function applied to data with duplicate labels </font>

## [2. `pivot_wider()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%" .smaller}

### Why would I pivot wider?

-   Some analyses require wide format data
-   For example, SEM in `lavaan` in R requires that both indicators and time are wide format.
-   The code below uses the codebook to rename items according to a numbered format that isn't specific to any trait

::: fragment
```{r}
big5 <- codebook |> filter(category == "big5")
soep_lavaan <- soep_long |>
  filter(category == "big5") |>
  mutate(item_name = mapvalues(item_name, big5$item_name, big5$lavaan_name, warn_missing = F))
soep_lavaan
```
:::

## [2. `pivot_wider()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%" .smaller}

### Why would I pivot wider?

-   Change the `soep_lavaan` data frame to be in wide format using `pivot_wider()`:
    -   pull the `names` from two sources: `item_name()` and `year`

::: fragment
```{r}
soep_lavaan |>
  pivot_wider(
    names_from = c("item_name", "year")
    , values_from = "value"
  )
```
:::

# dplyr: \_join() {background-color="purple"}

## [The `_join()` Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

::::: columns
::: {.column width="60%"}
-   Often we may need to pull different data from different sources
-   There are lots of reasons to need to do this
-   We don't have time to get into all the use cases here, so we'll talk about them in high level terms
:::

::: {.column width="40%"}
-   We'll focus on:
    -   `full_join()`
    -   `inner_join()`
    -   `left_join()`
    -   `right_join()`
:::
:::::

## [3. `full_join()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

:::::: columns
::: column
-   Most simply, we can put those back together keeping all observations.
-   Pro: sometimes we want to maintain missing data (i.e. some people are randomly missing variables and we don't want to drop them completely)
-   Con: can leave you with lots of `NA`s
:::

:::: column
-   Join the codebook to the data below using `full_join()`
-   Look at the data. What's going on here

::: fragment
```{r, eval = F}
soep_long |>
  filter(!category == "big5") |> 
  full_join(
    # your code here
  ) |> 
    View()
```
:::
::::
::::::

## [3. `full_join()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

\vspace{2em}

::: fragment
```{r}
soep_long |>
  filter(!category == "big5") |>
  full_join(codebook |> select(category, label, item_name, year, item_text)) 
```
:::

-   note we have lots of missing data because the Big Five portions of the codebook were joined even though we removed that data

## [4. `inner_join()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

:::::: columns
::: column
-   We can also keep all rows present in *both* data frames
-   Pro: Won't add rows with missing values in key variables
-   Con: will drop observations that you want want for counts, correlations, etc.
:::

:::: column
-   Join the codebook to the data below using `inner_join()`
-   Look at the data. What's going on here

::: fragment
```{r, eval = F}
soep_long |>
  filter(!category == "big5")
  full_join(
    # your code here
  ) |> 
    View()
```
:::
::::
::::::

## [4. `inner_join()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   Note that filtering, renaming/selecting, and joining is a common workflow

::: fragment
```{r}
soep_long |>
  filter(category == "big5") |>
  select(Proc_SID, trait = label, item_name, year, p_value = value) |>
  inner_join(
    soep_long |>
    filter(category == "le") |>
    select(Proc_SID, le = label, year, le_value = value)
  )
```
:::

## [5. `left_join()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   Or all rows present in the left (first) data frame, perhaps if it's a subset of people with complete data

::: fragment
```{r}
soep_long |>
  filter(category == "big5") |>
  select(Proc_SID, trait = label, item_name, year, p_value = value) |>
  left_join(
    soep_long |>
    filter(category == "le") |>
    select(Proc_SID, le = label, year, le_value = value)
  ) %>%
  print(n = 8)
```
:::

## [5. `right`_join()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

-   Or all rows present in the right (second) data frame, such as I do when I join a codebook with raw data

::: fragment
```{r}
soep_long |>
  filter(category == "big5") |>
  select(Proc_SID, trait = label, item_name, year, p_value = value) |>
  right_join(
    soep_long |>
    filter(category == "le") |>
    select(Proc_SID, le = label, year, le_value = value)
  ) %>%
  print(n = 8)
```
:::

## [Your Turn]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

In small groups, discuss what's happening when you use `full_join()`, `left_join()`, `right_join()`, `inner_join()`, and `anti_join()` with the code below. Which is correct in this use case?

```{r, eval = F, error = F}
soep_long |>
  filter(category == "big5") |>
  select(Proc_SID, trait = label, item_name, year, p_value = value) |>
  [x]_join(
    soep_long |>
    filter(category == "le") |>
    select(Proc_SID, le = label, year, le_value = value)
  )
```

# dplyr: split-apply-combine {background-color="#9B1003"}

## [Bringing it all together: Split-Apply-Combine]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

-   Much of the power of `dplyr` functions lay in the split-apply-combine method

-   A given set of of data are:

    -   *split* into smaller chunks
    -   then a function or series of functions are *applied* to each chunk
    -   and then the chunks are *combined* back together

## [3. `group_by()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

-   The `group_by()` function is the "split" of the method
-   It basically implicitly breaks the data set into chunks by whatever bare quoted column(s)/variable(s) are supplied as arguments.

## [4. `mutate()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

-   `mutate()` is one of your "apply" functions
-   When you use `mutate()`, the resulting data frame will have the same number of rows you started with
-   You are directly mutating the existing data frame, either modifying existing columns or creating new ones

## [5. `summarize()` / `summarise()`]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

-   `summarize()` is one of your "apply" functions
-   The resulting data frame will have the same number of rows as your grouping variable
-   You number of groups is 1 for ungrouped data frames

## [Exercise 1]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Remember when I said that long format data are easier to clean. Let's do that now.

## [Question 1]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Let's start with the Big Five data:

1.  `filter()` out only Big Five rows
2.  `mutate()` each observation so that values less than one are changed to `NA`
3.  Remove any missing values using `filter()` or `drop_na()`
4.  Group (split) the data so that you have a "group" for each person x trait x year combination
5.  `summarize()` the values to get a composite score for each Big Five trait for each person in each year:

## [Question 1: Solution]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

Remember when I said that long format data are easier to clean. Let's do that now.

```{r}
soep_big5 <- soep_long |>
  filter(category == "big5") |>
  mutate(value = mapvalues(value, seq(-8,0), rep(NA, 9))) |>
  drop_na(value) |>
  group_by(Proc_SID, label, year) |>
  summarize(value = mean(value)) |>
  ungroup()
soep_big5
```

## [Question 2]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%" .smaller}

Now let's take care of the life event data:\
1. `filter()` out only life event rows\
2. `mutate()` each observation so that

-   -2 = 0
-   1 = 1
-   everything else is `NA`

3.  Remove any missing values using `filter()` or `drop_na()`
4.  Group (split) the data so that you have a "group" for each person x event combination
5.  `summarize()` the values to get a `sum` score for each event for each person across all years:

## [Question 2: Solution]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

\vspace{2em}

```{r}
soep_le <- soep_long |>
  filter(category == "le") |>
  mutate(value = mapvalues(value, seq(-8,1), c(rep(NA, 6), 0, NA, NA, 1))) |>
  drop_na(value) |>
  group_by(Proc_SID, label) |>
  summarize(value = sum(value)) |>
  ungroup()
soep_le
```

## [Question 3]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

Just for practice, now make your Big Five data frame wide, leaving the time variable (year) long

## [Question 3: Solution]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

\vspace{2em}

::: fragment
```{r}
soep_big5 |>
  pivot_wider(
    names_from = "label"
    , values_from = "value"
  )
```
:::

## [Question 4]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Now, let's join the data frames back together.
-   Which join function do you think is most appropriate?
-   Hint: You will need to rename the `label` and `value` columns to reflect the category of the data

## [Question 4: Solution]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

\vspace{2em}

```{r}
soep_clean <- soep_big5 |>
  rename(trait = label, p_value = value) |>
  inner_join(
    soep_le |>
      rename(le = label, le_value = value)
  )
soep_clean
```

# Functions {background-color="#016295"}

## [Functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/mykonos-top.png" background-size="100%"}

**How to approach writing functions**? (*broad recipe*)

1.  Experiment with performing the task outside of a function
    -   Experiment with performing task with different sets of inputs
    -   Often, you must revise this code, when an approach that worked outside a function does not work within a function
2.  Write the function
3.  Test the function
    -   Try to "break" it
4.  **Continual improvement**. As you use the function, make continual improvements going back-and-forth between steps 1-3

## [Basics of writing functions]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/mykonos-top.png" background-size="100%" .smaller}

**Three components** of a function:

1.  **Function name**
    -   Define a function using `function()` and give it a **name** using the assignment operator `<-`
2.  **Function arguments** (sometimes called "inputs")
    -   Inputs that the function takes; they go inside the parentheses of `function()`
        -   Can be vectors, data frames, logical statements, strings, etc.
    -   In the above hypothetical code, the function took three inputs `arg1`, `arg2`, `arg3`, but we could have written:
        -   `function(x, y, z)` or `function(Larry, Curly, Moe)`
    -   In the "function call," you specify values to assign to these function arguments
3.  **Function body**
    -   What the function does to the inputs
    -   Function body goes inside the pair of curly brackets (`{}`) that follows `function()`
    -   Above hypothetical function doesn't do anything, but your function can **return a value** (covered in [later section](#return-values))

## [Exercise 2]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

Some common tasks when working with survey data:

-   Identify number of observations with `NA` values for a specific variable
-   Identify number of observations with negative values for a specific variable
-   Replace negative values with `NA` for a specific variable

## [Exercise 2: `num_negative()` function]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%" .smaller}

**Task**: Write function called `num_negative()`

-   Write a function that counts the number of observations with negative values for a specific variable
-   Apply this function to variables from dataframe `df` (created below)
-   Adapted from Ben Skinner's *Programming 1* R Workshop [HERE](https://www.btskinner.me/rworkshop/modules/programming_one.html)

```{r, echo=-c(1:3)}
set.seed(54321) # so that we all get the same random numbers
df <- tibble('id' = 1:100,
             'age' = sample(c(seq(11,20,1), -97,-98,-99),
                            size = 100,
                            replace = TRUE,
                            prob = c(rep(.09, 10), .1,.1,.1)),
             'sibage' = sample(c(seq(5,12,1), -97,-98,-99),
                               size = 100,
                               replace = TRUE,
                               prob = c(rep(.115, 8), .1,.1,.1)),
             'parage' = sample(c(seq(45,55,1), -4,-7,-8),
                               size = 100,
                               replace = TRUE,
                               prob = c(rep(.085, 11), .1,.1,.1))
)

# Sample dataframe `df` that contains some negative values
df
```

## [Steps:]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

**Recommended steps**:

-   [Perform task outside of function]{style="color:blue;"}
    -   HINT: `sum(data_frame_name$var_name<0)`
-   [Write function]{style="color:#9B1003;"}
-   [Apply/test function on variables]{style="color:purple;"}

## [Step 1: Perform task outside of function]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

\vspace{2em}

```{r}
names(df) # identify variable names
df$age # print observations for a variable

#BaseR
sum(df$age<0) # count number of obs w/ negative values for variable "age"
```

## [Step 2: Write function]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

\vspace{2em}

```{r}
num_missing <- function(x){
  sum(x<0)
}
```

## [Step 3: Apply function]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

\vspace{2em}

```{r}
num_missing(df$age)
num_missing(df$sibage)
```

## [Exercise 3: `num_missing()` function]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

In survey data, negative values often refer to reason for missing values:

-   E.g., `-8` refers to "didn't take survey"
-   E.g., `-7` refers to "took survey, but didn't answer this question"

## [Exercise 3: `num_missing()` function]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

**Task**: Write function called `num_negative()`

-   Write a function that counts number of missing observations for a variable and allows you to specify which values are associated with missing for that variable. This function will take two arguments:
    -   `x`: The variable (e.g., `df$sibage`)
    -   `miss_vals`: Vector of values you want to associate with "missing" variable
        -   Values to associate with missing for `df$age`: `-97,-98,-99`
        -   Values to associate with missing for `df$sibage`: `-97,-98,-99`
        -   Values to associate with missing for `df$parage`: `-4,-7,-8`

## [Steps]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

**Recommended steps**:

-   Perform task outside of function
    -   HINT: `sum(data_frame_name$var_name %in% c(-4,-5))`
-   Write function
-   Apply/test function on variables

## [Step 1: Perform task outside of function]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/blue-top.png" background-size="100%"}

\vspace{2em}

```{r}
sum(df$age %in% c(-97,-98,-99))
```

## [Step 2: Write function]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/red-top.png" background-size="100%"}

\vspace{2em}

```{r}
num_missing <- function(x, miss_vals){

  sum(x %in% miss_vals)
}
```

## [Step 3: Apply function]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/purple-top.png" background-size="100%"}

\vspace{2em}

```{r}
num_missing(df$age,c(-97,-98,-99))
num_missing(df$sibage,c(-97,-98,-99))
num_missing(df$parage,c(-4,-7,-8))
```

# `purrr` {background-color="#F3A51A"}

## [purrr::`map()`]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/yellow-top.png" background-size="100%"}

-   `map()` functions are the tidyverse alternative to for loops and chaotic lists with deep nesting structures
-   `map()` functions, unlike `_apply()` functions can take any number of inputs, which mimics nested `for` loops
-   `map()` functions can return any output type, including heterogeneous outputs (at least if you return it as a list)

## [purrr::`map()` inputs]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/yellow-top.png" background-size="100%"}

-   You control how many inputs using the following:
    -   `map()`: one input, arguments are `map(.x, .f)`
    -   `map2()`: two inputs, arguments are `map2(.x, y., .f)`
    -   `pmap()`: any number of inputs, arguments are `pmap(.l, .f)`
        -   Note the `.l` becuase this means we have to wrap inputs in a `list()`

## [purrr::`map()` outputs]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/yellow-top.png" background-size="100%"}

-   You can also control the output of `purrr::map()`:
    -   `map()`: outputs a list
    -   `map_chr()`: outputs a character vector\
    -   `map_dbl()`: outputs a numeric vector\
    -   `map_lgl()`: outputs a logical vector\
    -   `map_int()`: outputs a integer vector\
    -   `map_vec()`: outputs essentially any type of vector
-   Note that if one input combination fails, all will fail and nothing will be outputted

## [purrr::`map()` error handling]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/yellow-top.png" background-size="100%"}

-   Having everything fail because one thing went wrong is really frustrating
-   There are a number of functions in `purrr` to help with that:
    -   `possibly(.f, otherwise)`: returns whatever you ask it return with `otherwise` when a .f call fails
    -   `safely(.f)`: returns a list with the output, if successful, and errors, if unsuccessful
    -   Others: see [documentation](https://purrr.tidyverse.org/reference/index.html#adverbs).

## [purrr list columns]{style="color:black;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/yellow-top.png" background-size="100%" .smaller}

-   One of the easiest ways to work with `purrr` is using list columns in nested data frames

::::::: columns
:::: column
-   You can create a nested data frame using `tidyr::nest()` or `tibble()` (where one column is a list itself)

::: fragment
```{r}
soep_clean |>
  group_by(trait, year, le) |>
  nest() |>
  ungroup()
```
:::
::::

:::: column
-   You can then call `map()` within a mutate call to modify the list column or create new columns in your data frame

::: fragment
```{r}
tibble(
  x = c(1,2,3)
  , y = list(letters[1:5], letters[6:10], letters[11:15])
)
```
:::
::::
:::::::

# Exercise 4 {background-color="black"}

## [Question 1:]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Create a data frame called `soep_nested` that creates a list column of the data split by trait and life event.

::: fragment
```{r}
soep_nested <- soep_clean |>
  group_by(trait, le) |>
  nest() |>
  ungroup()
```
:::

## [Question 2:]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Using `mutate()`, create a new list column called `model` that runs the following function

::: fragment
```{r}
lmer_fun <- function(d){
  d <- d |> 
    mutate(wave = year - 2005) |>
    group_by(Proc_SID) |>
    filter(n() > 1)
  m <- lmer(p_value ~ wave + le_value + le_value:wave + (1 + wave | Proc_SID), data = d)
  return(m)
}
```
:::

## [Question 2: Solution]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

::: nonincremental
-   Using `mutate()`, create a new list column called `model` that runs the following function
:::

```{r}
lmer_fun <- function(d){
  d <- d |> 
    mutate(wave = year - 2005) |>
    group_by(Proc_SID) |>
    filter(n() > 1)
  m <- lmer(p_value ~ wave + le_value + le_value:wave + (1 + wave | Proc_SID), data = d)
  return(m)
}

soep_nested <- soep_nested |>
  mutate(model = map(data, lmer_fun))
soep_nested
```

## [Question 3:]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Use the following function to extract the number of people we estimated slopes for in this model. Output the result as an integer to a new column called `npeople`

::: fragment
```{r}
nslopes_fun <- function(m) summary(m)$ngrps
```
:::

::: fragment
```{r}
soep_nested <- soep_nested |>
  mutate(npeople = map_int(model, nslopes_fun))
soep_nested
```
:::

## [Question 4:]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Use the `tidy()` function from the `broom.mixed` package to extract the coefficients from the model and their confidence intervals. Save it to the column "tidy"
-   Hints:
    -   Use the argument `conf.int = T` to get the confidence intervals
    -   Additional arguments to the `.f` function called in `map()` can be just included as addition arguments (e.g., `map(.x, .f, conf.int = T)`)

## [Question 4: Solution]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

\vspace{2em}

```{r}
soep_nested <- soep_nested |>
  mutate(tidy = map(model, broom.mixed::tidy, conf.int = T))
```

## [Question 5:]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Let's practice making a super simple table. Do the following:

1.  remove the data and model columns from the data frame
2.  `unnest()` the `tidy` column
3.  Keep only fixed effects (`effect == "fixed"`)
4.  We only care about the interaction, so only keep the interaction `term`
5.  round the estimate, conf.low, and conf.high columns to 2 decimal places
6.  Keep the `trait`, `le`, `estimate`, `conf.low`, and `conf.high` columns only
7.  `pivot_wider()` by trait for `estimate`, `conf.low`, and `conf.high`

## [Question 5: Solution]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

\vspace{2em}

```{r}
soep_tab <- soep_nested |>
  select(-data, -model) |>
  unnest(tidy) |>
  filter(effect == "fixed" & grepl(":", term)) |>
  mutate(across(c(estimate, conf.low, conf.high), \(x) round(x, 2))) |>
  select(trait, le, estimate, conf.low, conf.high) |>
  pivot_wider(
    names_from = "trait"
    , names_glue = "{trait}_{.value}"
    , values_from = c(estimate, conf.low, conf.high)
  )
soep_tab %>% print(n = 6)
```

## [Question 6:]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

Use the function below to get model predictions

```{r}
pred_fun <- function(m){
  d <- m@frame |>
      select(-p_value) |>
      distinct()
  bind_cols(d, pred = predict(m, newdata = d))
}
```

::: fragment
```{r}
soep_nested <- soep_nested |>
  mutate(pred = map(model, pred_fun))
soep_nested
```
:::

## [Question 7:]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%"}

-   Let's practice making a super simple table. Do the following:

1.  remove the data, model, and tidy columns from the data frame
2.  `unnest()` the `tidy` column
3.  `group_by()` life event and `nest()` + `ungroup()`
4.  save this as `soep_pred`

## [Question 7: Solution]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

\vspace{2em}

```{r}
soep_pred <- soep_nested |>
  select(-data, -model, -tidy) |>
  unnest(pred) |>
  group_by(trait) |>
  nest() |>
  ungroup()
soep_pred
```

## [Question 8:]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/black-top.png" background-size="100%" .smaller}

-   Use the following function to create a new column `p` that contains spaghetti plots
-   Note that the function takes two inputs!

```{r}
spag_plot_fun <- function(d, trait){
  set.seed(6)
  d |>
    group_by(le) |>
    nest() |>
    mutate(data = map(data, ~filter(., Proc_SID %in% sample(unique(.$Proc_SID), 100)))) |>
    unnest(data) |>
    ungroup() |>
    mutate(le_value = ifelse(le_value > 1, 1, le_value)) |>
    ggplot(aes(x = wave, y = pred)) + 
      geom_line(aes(group = Proc_SID, color = factor(le_value)), alpha = .3) + 
      geom_smooth(method = "lm", se = F, color = "darkblue") + 
      scale_color_manual(values = c("grey", "blue"), labels = c("No Event", "Event")) + 
      labs(x = "Wave", y = "Predicted Trait Levels", color = "Life Event", title = trait) + 
      facet_wrap(~le) + 
      theme_classic() + 
      theme(legend.position = c(.7, .1))
}
```

## [Question 8: Solution]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

\vspace{2em}

```{r, fig.width=9}
soep_pred <- soep_pred |>
  mutate(p = map2(data, trait, spag_plot_fun))

soep_pred$p[[1]]
```

# Wrap-Up {background-color="#00674F"}

## [Wrap-Up]{style="color:white;"} {background-color="white" background-image="https://github.com/emoriebeck/psc203a-data-WQ26/raw/main/psc203a-data-WQ26/images/green-top.png" background-size="100%"}

-   Today's goal was to review the coding concepts we've used so far and ask you to apply them using a series of guided examples
-   The biggest takeaway I wanted you to have is *chaining*, or how you can use `tidyverse` functions in chains to accomplish a bunch of goals simultaneously
-   We cleaned, composited, and ran 50 models across thousands of people, including predictions and tables in less than 100 lines of code. Just doing the models, `tidy()`, and `predict()` parts of that alone would have been 150 lines of code and introduced huge opportunities for errors!
